---
title: "Project proposal"
author: "Team RARE"
output: github_document
---

```{r load-packages-and-data, message = FALSE}
library(tidyverse)
library(broom)
library(readxl)
install.packages("ggridges")
library(ggridges)
#GOLD assessment results for all EHS students
EHS_all_students <- read_excel("../data/ignore/Promise Early Education 2024-2025 Data- CLEAN.xlsx", 
    sheet = "EHS (birth-3)")

#GOLD assessment results for white vs. non-white EHS students
EHS_with_race <- read_excel("../data/ignore/Promise Early Education 2024-2025 Data- CLEAN.xlsx", 
    sheet = "EHS Race")

#GOLD assessment results for DLL vs non-DLL EHS students
EHS_Dual_Language <- read_excel("../data/ignore/Promise Early Education 2024-2025 Data- CLEAN.xlsx", 
    sheet = "EHS Eng SE")

#GOLD assessment results for multi-year vs first-year participating EHS students
EHS_multi_year <- read_excel("../data/ignore/Promise Early Education 2024-2025 Data- CLEAN.xlsx", 
    sheet = "EHS MY-NMY")
```

## 1. Introduction

Promise Early Education is a program designated for low-income or vulnerable families in the Lewiston-Auburn area. They provide education to income-eligible children who receive social services, such as the SNAP program, from the government. Along with educating children, they provide services to students with mental health and learning disabilities. Additionally, Promise also provides health and nutritional support for children. 

To assess the students, Promise uses the GOLD assessment system. The GOLD system tracks students' development over time in specific areas such as social-emotional, physical, cognitive, and language abilities. It tracks students on a widely held expectation (WHE) score to determine if they are below, meeting, or exceeding developmental expectations. The data collected is specifically targeted for students aged 0-5, and includes raw scores, scaled scores, and national norm comparison. This data is then utilized in the classroom to help teachers create lesson plans and give individualized attention, all to support the children’s development. 

For this project, our research question will explore children's development progress in the Promise Early Education program. We will analyze and evaluate the developmental gains of three key populations: Dual Language Learners (DLLs), students receiving special education services (those with an Individualized Education Program, or IEP), multi-year participants, and students of different racial/ethnic backgrounds. By comparing the progress of these subgroups against their peers, this research aims to illuminate how different student populations are supported in the classroom. Some questions we hope to investigate are: 

1. How does the progress of dual language learners differ from their peers?

2. What are the weakest/strongest skill sets of dual language learners compared to standard expectations (categorized by physical, social-emotional, literacy, etc.)

3. How does the progress across a year differ for a multi-year student compared to a first time student?

4. How does the proficiency in various categories differ by age range (B-1, 1-2, 2-3)?

5. How does growth for students compare by different seasons (fall to winter, winter to spring)? 

6. Are there differences in progress between students of different races?



## 2. Data

Because of the complex structure of the original dataset, we chose to clean and organize the data in Google Sheets rather than R to simplify the reformatting process. Specifically, we created new columns indicating “yes” or “no” for Dual Language Learners, race, and multi-year students. We also cleaned by consolidating data from all Early Head Start (EHS) programs into unified columns, because the original dataset had some categories arranged in columns and others in rows.


- `Category` = `which aspect is being tested` 
- `WHE Bottom` = `Bottom widely held expectation scores`
- `WHE Top` = `Top widely held expectation scores` 
- `Average` = `Average scores` 
- `# Children` = `Number of children being tested` 
- `# Below` = `Number of students below average`
- `% Below` = `Percent of students below average`
- `# Meeting` = `Number of students meeting expectations` 
- `% Meeting` = `Percentage of students meeting expectations` 
- ` # exceeding` = `Number of students exceeding expectations` 
- `% exceeding` = `The percentage of students exceeding expectations`
- `Time period` = `Fall, winter, or spring trimesters` 
- `Age range` = `The age range of the students` 
- `White` = `If the student is white or not`
- `Dual Language` = `If the student is a dual language learner or not` 
- `Multi-Year` = `If the student is a Multi-Year student or not` 


```{r dimensions-and-glimpse-all-students}

#GOLD assessment results for all EHS students
dim(EHS_all_students)
glimpse(EHS_all_students)

```
```{r dimension-and-glimpse-race}
#GOLD assessment results for white vs. non-white EHS students
dim(EHS_with_race)
glimpse(EHS_with_race)
```

```{r dimensions-and-glimpse-dll}
#GOLD assessment results for DLL vs non-DLL EHS students
dim(EHS_Dual_Language)
glimpse(EHS_Dual_Language)
```

```{r dimensions-and-glimpse-multi-year}
#GOLD assessment results for multi-year vs first-year participating EHS students
dim(EHS_multi_year)
glimpse(EHS_multi_year)
```

## 3. Data analysis plan

The variables we will be visualizing to explore the research questions include:

`Dual Language`

`category`

`Average`

`Multi-year`

X (TBD- waiting for IEP data)

`Time period` 

`White`

Types of graphs we may want to use:

- Line plot : to show the distribution of scores deviating away from the average for each `category` \n
- Waffle plots : To compare multi-year and not multi-year students \n
- Box Plot : showing the range of average scores\n
- Bar Graph : to compare students meeting, exceeding, and below expectations \n
- Bar Graph : to compare differences among distinctions like race and language \n


```{r preliminary-exploratory-data-analysis-bar-graph}
EHS_with_race |>
  mutate(Race_Group = if_else(`White (Yes or No)` == "Yes", "White", "Non-White")) |>
  group_by(Category, Race_Group) |>
  summarise(
    Weighted_Average = sum(Average * `# Children`) / sum(`# Children`),
    Total_Children = sum(`# Children`)
  ) |>
  ggplot(aes(x = Category, y = Weighted_Average, fill = Race_Group)) +
  geom_col(position = position_dodge()) +
  geom_text(
    aes(label = paste0("n=", Total_Children)),
    position = position_dodge(width = 0.9),
    vjust = -0.2
  ) +
  labs(
    title = "Weighted Average GOLD Assessment Scores by Category and Race",
    subtitle = "Counts (n) represent the total number of children contributing to each weighted average.",
    x = "Category",
    y = "Weighted Average Score",
    fill = "Race"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Our first preliminary exploratory data analysis is a bar graph comparing GOLD assessment scores between races using a weighted average. This visualization shows that generally, scores are similar between races, except for non-white students scoring significantly better in mathematics. Exploring this graph is useful as it showed us the importance of a weighted averages in our dataset. When we have different number of students, the average value of the score will turn out differently. It also shows us what visualizations work and don't work with data regarding race, for example, a ridge plot would not be able to show this data because we don't have enough average data points.

