---
title: "Project memo"
author: "Team Rare"
output: github_document
---

This document should contain a detailed account of the data clean up for your data and the design choices you are making for your plots. For instance you will want to document choices you've made that were intentional for your graphic, e.g. color you've chosen for the plot. Think of this document as a code script someone can follow to reproduce the data cleaning steps and graphics in your handout.

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
```

## Data Clean Up Steps for Overall Data

### Step 1: _________

```{r}

```

### Step 2: ________

## Plots

### Plot 1: Average Skill Scores: DLL (Dual Language Learners) vs. Non-DLL Students

```{r plot-1}
EHS_Dual_Language |>
  mutate(
    DLL_Group = if_else(`Dual Langage` == "Yes", "DLL", "Non-DLL"),
    `Time Period` = fct_relevel(`Time Period`, 
                                "Fall 2024/2025", 
                                "Winter 2024/2025", 
                                "Spring 2024/2025"),
    Average = as.numeric(Average)
  ) |>
  group_by(Category, DLL_Group, `Time Period`) |>
  summarise(
    Weighted_Average = sum(Average * `# Children`, na.rm = TRUE) / sum(`# Children`, na.rm = TRUE),
    Total_Children = sum(`# Children`, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = `Time Period`, y = Weighted_Average, 
             color = DLL_Group, group = DLL_Group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ Category) +
  labs(
    title = "Average Skill Scores: DLL (Dual Language Learners) vs. Non-DLL Students",
    caption = "Non-DLL students scored higher than DLL students in all skill areas, with both groups improving from Fall to Spring.",
    x = "Assessment Period (2024–2025 School Year)",
    y = "Weighted Average Score",
    color = "Group"
   ) + scale_x_discrete(
  labels = c(
    "Fall 2024/2025" = "Fall",
    "Winter 2024/2025" = "Winter",
    "Spring 2024/2025" = "Spring"
  )
) +
  scale_color_brewer(palette = "Set2") + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(t = 10, r = 10, b = 40, l = 10),
    plot.caption = element_text(hjust = 0.5),
  )
ggsave("plot_1.png")
```


```


#### Final Plot 1

```{r}

```


### Plot 2: _________

### Plot 3: ___________

Add more plot sections as needed. Each project should have at least 3 plots, but talk to me if you have fewer than 3.

### Plot 4: ___________ 

