---
title: "Promise Early Education"
author: "RARE"
output: github_document
---

```{r load-packages-and-data, message = FALSE}
library(tidyverse)
library(broom)
library(readxl)
install.packages("ggridges")
library(ggridges)
#GOLD assessment results for all EHS students
EHS_all_students <- read_excel("../data/ignore/Promise Early Education 2024-2025 Data- CLEAN.xlsx", 
    sheet = "EHS (birth-3)")

#GOLD assessment results for white vs. non-white EHS students
EHS_with_race <- read_excel("../data/ignore/Promise Early Education 2024-2025 Data- CLEAN.xlsx", 
    sheet = "EHS Race")

#GOLD assessment results for DLL vs non-DLL EHS students
EHS_Dual_Language <- read_excel("../data/ignore/Promise Early Education 2024-2025 Data- CLEAN.xlsx", 
    sheet = "EHS Eng SE")

#GOLD assessment results for multi-year vs first-year participating EHS students
EHS_multi_year <- read_excel("../data/ignore/Promise Early Education 2024-2025 Data- CLEAN.xlsx", 
    sheet = "EHS MY-NMY")
```

## Data Clean Up Steps for Overall Data

Since the data that we received from Promise Early education was not formatted well to put into R and would be extremely complicated to clean up using code, we did all of our data cleaning in Google sheets. This included formatting all variables into columns, and adding factors such as race, years participated in the program, language status, year, etc. for groups of students. 

## Plots

#### Final Plot 1: Comparing GOLD assessment scores of Dual Language vs. Non Dual-Language Learners 

```{r dual-language-plot, fig.alt = Faceted line chart of weighted average Dual Language Learners (DLL) and Non-DLL student scores across six skill areas (Cognitive, Language, Literacy, Mathematics, Physical, and Social Emotional) from Fall to Spring of the 2024–2025 school year.Across all categories, Non-DLL students score slightly higher than DLL students, with both groups showing consistent improvement from Fall to Spring. Lines are color-coded by group,DLL (green) and Non-DLL (orange).}
EHS_Dual_Language |>
  mutate(
    DLL_Group = if_else(`Dual Langage` == "Yes", "DLL", "Non-DLL"),
    `Time Period` = fct_relevel(`Time Period`, 
                                "Fall 2024/2025", 
                                "Winter 2024/2025", 
                                "Spring 2024/2025"),
    Average = as.numeric(Average)
  ) |>
  group_by(Category, DLL_Group, `Time Period`) |>
  summarise(
    Weighted_Average = sum(Average * `# Children`, na.rm = TRUE) / sum(`# Children`, na.rm = TRUE),
    Total_Children = sum(`# Children`, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = `Time Period`, y = Weighted_Average, 
             color = DLL_Group, group = DLL_Group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ Category) +
  labs(
    title = "Average Skill Scores: DLL (Dual Language Learners) vs. Non-DLL Students",
    caption = "Non-DLL students scored higher than DLL students in all skill areas, with both groups improving from Fall to Spring.",
    x = "Assessment Period (2024–2025 School Year)",
    y = "Weighted Average Score",
    color = "Group"
   ) + scale_x_discrete(
  labels = c(
    "Fall 2024/2025" = "Fall",
    "Winter 2024/2025" = "Winter",
    "Spring 2024/2025" = "Spring"
  )
) +
  scale_color_brewer(palette = "Set2") + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(t = 10, r = 10, b = 40, l = 10),
    plot.caption = element_text(hjust = 0.5),
  )
ggsave("dual-language-plot.png")
```



### Final Plot 2: Comparing GOLD assessment scores of Multi-Year vs. First Time participants of the Promise Program 

Using an accessible color palette, we looked at the GOLD assessment scores for multi-year vs first time participants of the Promise program.  We looked at the weighted average of the scores for each category. For First-Year participants, the sample size is 30 children, and for Multi-Year participants, the sample size is  51 children. We organized the bar graph with the categories on the axis, and scores on the x-axis, and ordered them based on the highest weighted average score (from top down). The graph shows higher scores in each category for multi-year students, proving the effectiveness of the Promise curriculum. 

```{r multi-year-plot, fig.alt = Bar Chart of GOLD Assessment Scores for Multi-Year Students: Compares the weighted average scores of First-Year, n=30 students, and Multi-Year, n=51 students, participants across six developmental categories. The chart shows that Multi-Year students achieved higher average scores in all six developmental areas, indicating greater mastery of developmental objectives.}
EHS_multi_year |>
  mutate(
    Program_Status = if_else(`Multi-Year` == "yes", "Multi-Year", "First-Year")) |>
  group_by(Category, Program_Status) |>
  summarise(
    Weighted_Average = sum(Average * `# Children`) / sum(`# Children`),
    Total_Children = sum(`# Children`),
    .groups = 'drop'
  ) |>
  mutate(
    Legend_Label = paste0(Program_Status, " (n=", Total_Children, ")")
  ) |>
  mutate(
    Category = forcats::fct_reorder(Category, Weighted_Average, .fun = max)
  ) |>
  ggplot(aes(x = Category, y = Weighted_Average, fill = Legend_Label)) +
  geom_col(position = position_dodge()) +
  labs(
    title = "Higher Average Gold Assessment Scores for Multi-Year Students",
    subtitle = "Categories ranked top-down by the highest weighted average score", 
    y = "Score", 
    x = " ",
    fill = "Program Status (Total Children)",
    caption = "Data reflects scores weighted by child enrollment (n) included in the legend. \n Higher scores on the GOLD assessment indicate greater mastery of developmental objectives."
  ) +
  coord_flip() +
  scale_fill_brewer(palette = "Dark2") +
    theme(
    plot.caption = element_text(
      size = 9, 
      hjust = 0.5,
      face = "italic"
    )
  )
ggsave("multi-year-plot.png")
```


### Final Plot 3: Comparing progress of GOLD assessment scores from different ages of the Early Head Start Program (EHS)

We initially started with a specific color palate but changed the colors to a more accessible color pattern. Furthermore, we reordered the bars in each bar graph in numerical order. In these graphs, we focused on how different age groups impacted the weighted average assessment scores for GOLD across various categories. Mathematics, regardless of age, consistently scored the lowest while Physical scored the highest. This relatively similar trend occurred over all three tested age ranges, with only the average values increasing. The graphs were also focused on the time period between three seasons: Fall 2024/2025, Winter 2024/2025, and Spring 2024/2025. 


``` {r agerange1to2, fig.alt = "Vertical bar chart looking at the five different categories (Physical, Cognitive, Language, Social-Emotional, and Mathematics) and the weighted average per category for ages 1 to 2 showing how age changes progress. display weighted averages of roughly 430, 380, 330, 325. All bars appear same color (gold)."}
age_one <- EHS_all_students |>
  filter(`age range` == "1-2") |>
  group_by(Category) |>
  summarise(
    Weighted_Average = weighted.mean(Average, `# Children`),
    Children_N = sum(`# Children`),
    .groups = "drop"
  )

ggplot(age_one, aes(x = reorder(Category, Weighted_Average), y = Weighted_Average)) +
  geom_col(fill = "#E69F00") +
  labs(
    title = "Proficiency in Categories by Age Range",
    subtitle = "From Year 1 to Year 2",
    x = "Skills by Category", 
    y = "Weighted Average"
  )
ggsave("ageplot1.png")
```

``` {r agerange2to3, fig.alt = "Vertical bar chart looking at the five different categories (Physical, Cognitive, Language, Social-Emotional, and Mathematics) and the weighted average per category for ages 2 to 3 showing how age changes progress. display weighted averages of roughly 495, 415, 395, 370, and 270 respectively. All bars appear same color (blue)."}
age_two <- EHS_all_students |>
  filter(`age range` == "2-3") |>
  group_by(Category) |>
  summarise(
    Weighted_Average = weighted.mean(Average, `# Children`),
    Children_N = sum(`# Children`),
    .groups = "drop"
  )

ggplot(age_two, aes(x = reorder(Category, Weighted_Average), y = Weighted_Average)) +
  geom_col(fill = "#56B4E9") +
  labs(
    title = "Proficiency in Categories by Age Range",
    subtitle = "From Year 2 to Year 3",
    y = "Weighted Average", 
    x = "Skills by Category"
  )
ggsave("ageplot2.png")
```

``` {r agerangebto1, fig.alt = "Vertical bar chart looking at the five different categories (Physical, Cognitive, Language, Social-Emotional, and Mathematics) and the weighted average per category for ages birth year to 1 showing how age changes progress. display weighted averages of roughly 50, 200, 200, 215, and 300 respectively. All bars appear same color (green)."}
age_birth <- EHS_all_students |>
  filter(`age range` == "B-1") |>
  group_by(Category) |>
  summarise(
    Weighted_Average = weighted.mean(Average, `# Children`),
    Children_N = sum(`# Children`),
    .groups = "drop"
  )

ggplot(age_birth, aes(x = reorder(Category, Weighted_Average), y = Weighted_Average)) +
  geom_col(fill = "#009E73") +
  labs(
    title = "Proficiency in Categories by Age Range",
    subtitle = "From Birth to Year 1",
    y = "Weighted Average", 
     x = "Skills by Category"
  ) 
ggsave("ageplotbirth.png")
```


### Final Plot 4: Comparing GOLD assessment scores of students in the EHS program based on race

```{r race-plot}
EHS_with_race |>
  mutate(Race_Group = if_else(`White (Yes or No)` == "Yes", "White", "Non-White")) |>
  group_by(Race_Group) |>
  mutate(Overall_n = sum(`# Children`)) |>
  ungroup() |>
  group_by(Category, Race_Group) |>
  summarise(
    Overall_n = sum(`# Children`),
    Weighted_Average = sum(Average * `# Children`) / sum(`# Children`),
    Total_Children = sum(`# Children`)
  ) |>
  mutate(Legend_Label = paste0(Race_Group, " (n=", Overall_n, ")"))|>
  ggplot(aes(x = Category, y = Weighted_Average, fill = Legend_Label)) +
  geom_col(position = position_dodge()) +
  scale_fill_viridis_d()+
  labs(
    title = "Weighted Average GOLD Assessment Scores by Category and Race",
    subtitle = "Counts (n) represent the total number of children contributing to each weighted average.",
    x = "Category",
    y = "Weighted Average Score",
    fill = "Race"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("raceplot.png")
```
